# Migrating from Legacy Workflows {#sec-legacy}

Most public health data work predates the tools in this book. State health departments run surveillance on SAS programs written over twenty years by people who have since retired, weekly reports come out of Excel workbooks where the analysis lives in formula chains and pivot tables, and case data sits in Access databases on a shared drive. These workflows produce real, trusted outputs every week. The concern is whether the team can continue to maintain and staff them.

Translating syntax is only part of a migration. Even a familiar operation such as `PROC MEANS` requires checking how the R equivalent handles grouping and missing values. What takes judgment is deciding what to migrate and in what order, proving that the new pipeline produces the same numbers as the old one, and retiring the old system without losing the institutional knowledge embedded in it. This chapter is about that process.

## Why Migrate, and Why Not {#sec-legacy-why}

The case for migrating usually starts with money: SAS licensing is a recurring line item that R eliminates, and for a small team the savings alone can fund a training program. The hiring pool tilts the same direction. New analysts arrive knowing R or Python, the people who know SAS are retiring, and every year a workflow stays in SAS it gets harder to find someone who can maintain it. Beyond cost and staffing, an R workflow can integrate with the tools covered earlier: parameterized Quarto reports (@sec-reporting), dashboards (@sec-dashboards), scheduled pipelines on GitHub Actions (@sec-automation), or a version-controlled analysis that can be reviewed and handed off (@sec-git).

There are equally good reasons to leave a particular workflow alone, at least for now. A report that is being retired should not be rewritten; use migration planning to identify products that can be retired. A rewrite the team cannot maintain is worse than the status quo: if the only SAS analyst retires in six months and nobody else knows R yet, the first investment is training (@sec-team-upskilling). Timing matters too, since a rewrite in the middle of an outbreak response adds risk exactly when the team can least absorb it.

Account for the risk that a migration rewrites software that works. The legacy program's numbers are trusted; the rewrite's are not until you prove they match. Plan time for those comparisons.

## Inventory and Triage {#sec-legacy-triage}

Start by listing every recurring product the legacy stack produces: each report, each dataset handed to another program, each dashboard feed. For each one, record what runs it, how often, who consumes the output, and who understands the code. This inventory is usually revealing on its own. Most teams find a handful of products that matter enormously, a long tail that runs out of habit, and at least one program nobody can explain.

Then triage. Good first candidates for migration score high on all of these:

- It runs on a schedule (weekly, monthly), so the payoff recurs and there are regular opportunities to compare old against new.
- Someone who understands the current logic is still available to answer questions.
- The output has an engaged consumer who will notice, and care, if numbers change.
- It is painful today: manual steps, fragile hand-offs, a single person who can run it.

One-off historical analyses generally should not be migrated at all. Archive the code and its outputs, document what they were for, and leave them alone.

Migrate one product end to end before starting a second. A complete pilot, from raw data to delivered report, running in parallel with the legacy version, teaches the team more than partially migrating ten programs, and it gives stakeholders a working example to assess.

## Reading Legacy Data Formats {#sec-legacy-data}

Whatever else migrates, the data has to. The [`haven`](https://haven.tidyverse.org/) package reads SAS (`.sas7bdat`, `.xpt`), SPSS (`.sav`), and Stata (`.dta`) files directly:

```{r}
#| label: haven-read
#| message: false
library(haven)

path <- system.file("examples", "iris.sas7bdat", package = "haven")
read_sas(path)
```

Check labels, missing values, and dates during import. SAS stores value labels (formats) separately from the data, in a catalog file typically named `formats.sas7bcat`; pass it to `read_sas()` and the labels arrive as labelled columns, which `as_factor()` converts to R factors:

```{r}
#| label: haven-catalog
#| eval: false
cases <- read_sas(
  "data/raw/cases.sas7bdat",
  catalog_file = "data/raw/formats.sas7bcat"
)
cases <- as_factor(cases)
```

Missing values need more care. SAS distinguishes 28 numeric missing values (`.`, `._`, and `.A` through `.Z`), often used to encode *why* a value is missing: refused, unknown, not applicable. R has one `NA`. `haven` preserves the distinctions as tagged NAs (see `?haven::tagged_na`), but most downstream R code will treat them all as plain `NA`, so if the distinctions carry meaning, convert them to an explicit column before they are lost.

Dates come through correctly, because `haven` knows that SAS counts days from 1960-01-01 while R counts from 1970-01-01. Hand-rolled imports do not get that conversion: a CSV exported from SAS with raw date numbers will be off by exactly 3,653 days, an error distinctive enough to recognize on sight.

For Excel sources, `readxl` reads `.xlsx` and `.xls` files, and the spreadsheet-hygiene guidance in @sec-dataorg applies doubly to inherited workbooks. For Access, the `odbc` package can connect directly on Windows, where the Access ODBC driver is usually already installed:

```{r}
#| label: access-connect
#| eval: false
library(DBI)

con <- dbConnect(
  odbc::odbc(),
  .connection_string = paste0(
    "Driver={Microsoft Access Driver (*.mdb, *.accdb)};",
    "Dbq=S:/surveillance/cases.accdb;"
  )
)
dbListTables(con)
```

Whichever format the data arrives in, convert it once and save an analysis-ready copy in an open format (CSV or Parquet, per @sec-dataorg), keeping the original file as the untouched raw archive. Proprietary formats depend on software that will not always be there to read them.

## Translating SAS Code {#sec-legacy-translate}

SAS and R differ more in mindset than in capability. A SAS program is a sequence of DATA steps (implicit row-by-row loops that read one dataset and write another) punctuated by PROC calls. Idiomatic R is vectorized: operations apply to whole columns at once, chained with pipes. A literal line-by-line translation produces R code that works but reads like SAS; the better approach is to translate what each step accomplishes.

Common operations have R counterparts, though their defaults and edge cases can differ:

| SAS | R equivalent |
|---|---|
| DATA step assignments, `IF`/`THEN` | `mutate()` with `if_else()` or `case_when()` |
| `PROC SORT` (and `NODUPKEY`) | `arrange()`, `distinct()` |
| `PROC MEANS`, `PROC SUMMARY` | `group_by()` + `summarize()` |
| `PROC FREQ` | `count()`, `janitor::tabyl()` |
| `PROC TRANSPOSE` | `tidyr::pivot_longer()`, `pivot_wider()` |
| `MERGE` in a DATA step | Joins only after checking BY-key multiplicity and row-matching semantics |
| `PROC SQL` | dplyr, or SQL directly via DBI (@sec-databases) |
| `PROC FORMAT` | factor levels and labels |
| `PROC EXPORT` | `readr::write_csv()` |
| `%MACRO` | a function |

Macros require a different approach. SAS macros are text substitution; R functions evaluate arguments and return values, and they are easier to test and document. A macro library that grew over years often translates into a small set of R functions, which is the point at which a team package (@sec-package-development) starts to make sense.

### Check merge and missing-value semantics

A SAS DATA-step match-merge is not a SQL join. If one BY group has two records on the left and three on the right, a match-merge produces three records; an all-matches join produces six. Check duplicate keys, row order, retained values, and unmatched records before choosing an R join. The [SAS combining-data documentation](https://support.sas.com/documentation/cdl/en/lrcon/65287/HTML/default/n1tgk0uanvisvon1r26lc036k0w7.htm) explains the match-merge behavior.

[SAS numeric missing values compare below ordinary numbers](https://support.sas.com/documentation/cdl/en/lrcon/62955/HTML/default/a000989180.htm). R missing-value comparisons return `NA`. For example, SAS evaluates `count < 0` as true for `.` and `.A`, as well as for a negative count. This small check makes that difference explicit:

```{r}
#| label: legacy-missing-parity
count <- c(NA_real_, haven::tagged_na("a"), -1, 0, 5)
sas_expected <- c(TRUE, TRUE, TRUE, FALSE, FALSE)
r_comparison <- count < 0
sas_comparison <- is.na(count) | count < 0
stopifnot(identical(sas_comparison, sas_expected))
data.frame(count, r_comparison, sas_comparison)
```

This reproduces that particular SAS expression; it does not define the rule your analysis should use. If missing counts should be excluded, write and test that requirement in both versions. Tagged missing values preserve distinctions during import, but do not give R SAS's comparison semantics. Other operations need their own parity cases.

### AI-assisted translation

Legacy translation is a task AI coding assistants (@sec-ai) handle well. Claude Code (@sec-ai-claude-code) can take a several-hundred-line SAS program and produce a credible R draft in minutes, and it is equally useful in the other direction: asking it to *explain* an inherited program, step by step, before you translate anything. When the original author is gone, an AI walkthrough of what a complex DATA step does can help you investigate it.

Treat the output as a draft, though. Allow time for verification. AI translation can produce code that looks right and runs clean but handles an edge case (usually missing values) differently than the original. Nothing that comes out of an assistant should reach production without passing the parity checks in the next section. Check code as well as data before sharing it: scripts can contain credentials, internal paths, and sensitive values. The privacy considerations in @sec-ai-models apply to both. Confirm this use is permitted under your agency's AI policy before you start, and note that an assistant may read data files in your project directory (@sec-aigovernance-phi).

## Verifying Parity {#sec-legacy-parity}

The legacy program's output is your test oracle. Freeze an input dataset, run it through both pipelines, and compare the outputs with code. The [`waldo`](https://waldo.r-lib.org/) package gives readable, precise diffs:

```{r}
#| label: waldo-compare
sas_out <- data.frame(
  county = c("Fairfax", "Arlington", "Loudoun"),
  cases = c(23L, 41L, 18L),
  rate = c(2.0, 17.2, 4.3)
)

r_out <- data.frame(
  county = c("Fairfax", "Arlington", "Loudoun"),
  cases = c(23L, 41L, 18L),
  rate = c(2.0, 17.2, 4.2)
)

waldo::compare(sas_out, r_out)
```

The diff points straight at the discrepancy. Counts should match exactly; derived numbers like rates should match within a small tolerance (`waldo::compare()` takes a `tolerance` argument), because the two engines can legitimately differ in the last decimal places of floating-point arithmetic.

When outputs differ, the cause is usually a place where SAS and R behave differently by design. Check these differences:

| Behavior | SAS | R |
|---|---|---|
| Rounding halves | Away from zero: `ROUND(2.5)` is 3 | Half to even: `round(2.5)` is 2 |
| Missing values | `.`, `._`, and `.A`–`.Z` | A single `NA` |
| Missing values in sorts | Sort first (missing is smallest) | `arrange()` puts `NA` last |
| Blank strings | `""` is missing | `""` is a value, distinct from `NA` |
| Date origin | 1960-01-01 | 1970-01-01 |

The rounding difference is worth seeing once, because it looks so much like a bug:

```{r}
#| label: rounding
round(c(0.5, 1.5, 2.5))
```

R follows the IEEE standard and rounds halves to the nearest even digit, so 0.5 rounds to 0 and 2.5 rounds to 2, where SAS would give 1 and 3. If the legacy program rounds before a suppression threshold or publishes rounded rates, the two pipelines will disagree by one in the last digit on exactly the values ending in five, and only those.

Every discrepancy gets one of two resolutions: fix the R code to match, or accept the difference and write down why. Keep the comparison script and a short parity report in the repository alongside the migrated code, listing each accepted difference and its justification. That document is what lets you tell a program manager, with evidence, that the new numbers are trustworthy, and it is exactly the kind of artifact analytical peer review (@sec-peerreview) should examine.

## Running in Parallel and Cutting Over {#sec-legacy-cutover}

Parity on a frozen test dataset is necessary but not sufficient, because live data keeps finding cases the test data did not contain. So run both pipelines in parallel on real data for a fixed number of cycles: four to eight weeks for a weekly report is typical. Each cycle, produce both outputs, run the comparison script, and investigate anything new. These investigations are where the legacy program's undocumented behavior surfaces: the special-casing of one bad data year, the county recode nobody mentioned. Each one either becomes deliberate logic in the new pipeline or a documented, accepted difference.

Agree on the cutover criteria before the parallel period starts (for example: N consecutive cycles with no unexplained discrepancies), so the team can assess readiness against agreed criteria. Then, at cutover:

- Move the legacy code into the repository under a `legacy/` directory, read-only, with a README mapping each old program to its replacement (@sec-documentation-readme).
- Run the legacy pipeline one final time and archive its outputs as the last reference point.
- Actually decommission it. Remove it from the schedule, and cancel the license when the last workflow that needs it is gone.

The decommissioning step is the one teams skip, and skipping it is costly. Two live pipelines mean two sources of truth, and they will eventually diverge, at which point someone has to figure out which one the program has been using. Cancel unused licenses after confirming that no retained workflow needs them.

## Excel and Access Workflows {#sec-legacy-excel}

Spreadsheet migrations differ from SAS migrations because there is no program to translate. The analysis may be spread across formula chains, pivot caches, and cells someone updates by hand, which can be difficult to compare and review: one inserted row can silently break a range reference three sheets away.

The pattern that works is to split the workbook's two jobs. If data entry happens in Excel, let it keep happening there, in a workbook restructured to the rectangular, one-thing-per-cell standard of @sec-dataorg. Everything after entry moves to a script: the workbook becomes an *input* that `readxl` reads, and R produces the outputs. The translations are mostly one-to-one. A `VLOOKUP` chain is a `left_join()`, a pivot table is `group_by()` plus `summarize()` (with `pivot_wider()` for the layout), and logic encoded in conditional formatting becomes an explicit column. Validate imported data (@sec-validation) to detect entry errors before the analysis runs.

Access databases play two roles that migrate separately. As storage, the tables can be read directly via `odbc` as shown above, or exported once if the database is being retired. Simple select and aggregation queries may translate directly. [Microsoft documents differences between Access SQL and ANSI SQL](https://support.microsoft.com/en-us/access/comparison-of-microsoft-access-sql-and-ansi-sql). Inventory Access-specific functions, date syntax, crosstab queries, query parameters supplied by forms, and VBA dependencies separately; these can require substantial changes. Validate the translated query outputs before retiring the original application. If several people still need to enter and look up cases interactively, Access can keep that front-end job for now, with R connecting read-only for analysis; the longer-term home for shared, structured case data is an IT-managed database, which is a conversation to start with the people in @sec-governance-access.

## The People Side {#sec-legacy-people}

Legacy programs often contain undocumented analytical decisions. Somewhere in that SAS code is the exclusion rule from a 2015 data quality incident and the workaround for one lab's misformatted files, none of it written down anywhere else. The person who wrote the program holds the other half of that memory.

Involve the legacy author in the migration. Pair them with the team's strongest R programmer: the veteran explains what each step is for and reviews parity discrepancies, while the R partner writes idiomatic new code, and both learn. This is upskilling (@sec-team-upskilling) with a concrete deliverable attached, and it gives the author an opportunity to learn R, which can support retention (@sec-team-retention). When the author is already gone, the code and existing documentation may be your main sources, which makes the careful reading, and the AI-assisted explanation described earlier, that much more important.

Finally, pick the timing like the project decision it is (@sec-pm). A migration has a long middle where both systems run and the team carries double work. Schedule the parallel runs for a period with fewer reporting demands.

## Further Reading {#sec-legacy-reading}

- The [haven documentation](https://haven.tidyverse.org/) covers the details of reading SAS, SPSS, and Stata files, including labelled values and tagged missings.
- [*R for Excel Users*](https://rstudio-conf-2020.github.io/r-for-excel/) (Lowndes and Horst) is a free, workshop-format introduction to R aimed specifically at people whose current workflow is Excel.
- [*The Epidemiologist R Handbook*](https://epirhandbook.com/) includes a "Transition to R" chapter with practical advice, and R equivalents for common tasks, aimed at epidemiologists coming from SAS, Stata, and SPSS.
