# Package Loading {#sec-packageloading}

Loading packages in R is easy enough to do carelessly. A script accumulates ten or fifteen `library()` calls, and everything runs until it does not. An analyst's `select()` call starts returning the wrong results, and nothing errors to say so. Scrolling to the top of the script reveals that MASS was loaded after dplyr, giving `MASS::select()` precedence over `dplyr::select()`. No warning, just the wrong function running.

## How R's search path works {#sec-packageloading-searchpath}

When you call `library()`, R attaches the package to the *search path*, an ordered list of environments R searches when resolving a name. You can inspect it at any time:

```{r}
search()
```

After loading a few packages, the path grows:

```{r}
library(dplyr)
library(MASS)
```

`package:MASS` now sits above `package:dplyr` in the search path. When R encounters the name `select`, it finds `MASS::select` first and stops. R prints a message when this happens:

```r
# Attaching package: 'MASS'
# The following object is masked from 'package:dplyr':
#     select
```

In a script that loads fifteen packages, these messages scroll past. The script runs, and there is no error to flag that `select()` is now going to the wrong package.

::: {.callout-warning}
## R's masking message is easy to miss

R prints a note at load time when a package masks a function from an earlier-loaded one, but in a long block of `library()` calls the messages are easily overlooked. R replaces the earlier function without any error. The `conflicted` package, covered in @sec-packageloading-conflicted, changes this: any ambiguous name throws an error rather than silently resolving to whichever package was loaded last.
:::

## Load only what you need {#sec-packageloading-minimal}

Every `library()` call extends the search path. A package block loaded by habit rather than intention is a reliable source of masking bugs in shared scripts.

Load a package with `library()` only if you call its functions throughout the script using bare names. If a package appears once or twice, use `::` instead (see @sec-packageloading-doublecolon).

Loaded by habit:

```r
library(tidyverse)
library(lubridate)
library(janitor)
library(readxl)
library(openxlsx)
library(scales)
library(zoo)
library(forecast)
library(MASS)
library(car)
```

Trimmed to what the script uses:

```r
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
```

The shorter list documents dependencies. A reader can see what the script uses and trust the list is intentional.

::: {.callout-tip}
## Loading tidyverse

`library(tidyverse)` loads ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats, and lubridate in a single call. That is a reasonable choice when you are genuinely using several of those packages. For a script that only needs dplyr and ggplot2, loading individual packages is more explicit about what you depend on and puts fewer packages on the search path.

You'll get the following message when you load tidyverse:

```
── Attaching core tidyverse packages ──── tidyverse 2.0.0 ──
✔ dplyr     1.2.0     ✔ readr     2.1.6
✔ forcats   1.0.1     ✔ stringr   1.6.0
✔ ggplot2   4.0.1     ✔ tibble    3.3.0
✔ lubridate 1.9.4     ✔ tidyr     1.3.1
✔ purrr     1.2.0     
── Conflicts ────────────────────── tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
ℹ Use the conflicted package to force all conflicts to become errors
```
:::

## Use `::` for one-off functions {#sec-packageloading-doublecolon}

The `::` operator calls a function directly from a specific package without attaching it to the search path, so no masking occurs.

```r
# Reading a single Excel file
data <- readxl::read_excel("surveillance_data.xlsx")

# Reading a large CSV with data.table's fast parser
raw <- data.table::fread("large_file.csv")

# Fitting a distribution without loading MASS
fit <- MASS::fitdistr(x, "normal")
```

`::` also documents function origin at the call site. `readxl::read_excel()` names its source; `read_excel()` does not.

When to use `::` versus `library()`:

- Use `library()` when you call functions from a package throughout the script. Loading `dplyr` makes sense if you are writing a dozen `filter()` and `mutate()` calls.
- Use `::` when you need one or two functions from a package and do not use it elsewhere.

::: {.callout-note}
## `::` works even without `library()`

You do not need to call `library()` before using `::`. As long as a package is installed, `readxl::read_excel()` will find and call it. You get the function without attaching the package or affecting the search path.
:::

## The `conflicted` package {#sec-packageloading-conflicted}

The `conflicted` package [@conflicted2023] changes the default behavior: any ambiguous name throws an error, forcing an explicit choice rather than silently inheriting load order.

Install from CRAN:

```r
install.packages("conflicted")
```

Without `conflicted`, load order silently determines which `select()` runs:

```{r}
#| error: true
library(dplyr)
library(MASS)

# Silently uses MASS::select(), not dplyr::select()
result <- mtcars |> select(mpg, cyl)
```

With `conflicted` loaded first, the same call throws an informative error:

```{r}
#| error: true
library(conflicted)
library(dplyr)
library(MASS)

result <- mtcars |> select(mpg, cyl)
```

This is the point. Load order was deciding which `select()` to use; `conflicted` makes you decide.

Declare your preferences with `conflicts_prefer()`, once at the top of the script alongside your `library()` calls:

```{r}
library(conflicted)
library(dplyr)
library(MASS)

conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)

# Now select() and filter() unambiguously mean dplyr
result <- mtcars |> select(mpg, cyl)
```

Use `conflict_scout()` to see every conflict in your session before any function calls:

```{r}
conflict_scout()
```

Run it after loading your packages to see what needs a preference declared.

::: {.callout-tip}
## `conflicted` works well alongside tidyverse

Load `conflicted` before `tidyverse`, then declare preferences for the most common conflicts:

```{r}
library(conflicted)
library(tidyverse)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::lag)
```

After these declarations, `filter()`, `select()`, and `lag()` unambiguously refer to dplyr. Any other conflict produces an error rather than a silent wrong answer.
:::

## Team implications {#sec-packageloading-team}

Package load order affects reproducibility across the team. The same script can produce different results on different machines when analysts have different packages installed or load them in a different order. `dplyr::select()` always means `dplyr::select()`, but `select()` depends on who loaded what and when.

Code review (see @sec-peerreview) is a natural checkpoint for package hygiene. Reviewers should flag unnecessary `library()` calls and bare function names that commonly conflict. Both are easier to catch in review than when writing.

`renv` (see @sec-environments-renv) ensures everyone is using the same package versions, but it does not determine which `select()` runs when both dplyr and MASS are loaded. `renv` and `conflicted` are complementary: one locks down versions, the other makes conflict resolution explicit.

The package block at the top of a shared script documents what the script depends on. A bloated list loaded by habit, combined with silent masking, means different analysts running the same script may not be running the same code.
